\vspace{-0.8cm}Vertex Coloring and Applications

### 2017-05-11 work/paper.tex
Making almost final changes to paper

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{graphicx}
\usepackage{float}
\usepackage{titling}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[numbib,nottoc]{tocbibind}
\usepackage{enumitem}
\usepackage{multicol}
\setlist[itemize]{topsep=0pt}
\setlist[enumerate]{topsep=0pt}
\setlength\parindent{0pt}
\setlength{\droptitle}{-1.5cm}

@@

\title{Vertex Coloring and Applications}
\date{May 5, 2017}
\author{Shawn Seymour\\ University of Minnesota Morris}

@@

\section{Introduction}
Consider the map of the 48 contiguous states in the USA. Suppose we would like to color each state such that no two states that share a boundary have the same color. How many colors would it take to do this? This problem can be modeled as a \emph{graph}. In general, we could represent every state with a \emph{vertex} and draw an \emph{edge} between two states that share a border. A graph, denoted $G = (V, E)$, is a set of vertices $V$ and a set of edges $E$. A \emph{simple graph} is a loopless, undirected graph where no two edges connect the same pair of vertices. For our purpose, assume all graphs are simple graphs. This problem, along with many others, can be solved using vertex coloring.

\begin{figure}[H]
\centering
\includegraphics[width=0.7\linewidth]{figures/map-colors.pdf}
\caption{A proper coloring of the USA}\label{fig:map}
\end{figure}

@@

The \emph{chromatic number}, denoted $\chi(G)$, is the minimum number of colors needed to have a proper vertex coloring of a graph $G$. The \emph{vertex coloring problem} (VCP), when given a graph $G$, is to find $\chi(G)$. By applying the vertex coloring problem to our map example, we can determine how many colors one would need to color it with regards to our constraint. An example coloring of our map is shown in Figure \ref{fig:map}. The \emph{k-Colorability problem} asks if a graph can be colored with $k$ colors. If a graph can be colored with $k$ colors, it is said to be \emph{k-colorable}.

@@

\subsection{Classifying the Vertex Coloring Problem}
We call any problem that can be answered with ``yes'' or ``no'' a \emph{decision problem}. To classify the vertex coloring problem, we must first introduce some classes, defined in \cite{moret}:

@@

Take note that these classes are not exclusive: any member of P is also a member of NP. NP-complete problems can be thought of as the problems that are both in NP and NP-hard. Although it may seem like all problems that are NP are also NP-complete, this is not the case. There are a select few problems in NP that are not NP-complete or in P \cite{ladner1975structure}.

@@

It is NP-hard to compute the chromatic number of a graph. This can be proven by showing it is at least as hard as a problem in NP. This is typically done by a reduction (a transformation) to or from 3-SAT, which we'll examine later. This is proven in \cite{garey}.

@@

To prove this, we need to reduce a known NP-complete problem to the vertex coloring problem in polynomial time. By reducing a known NP-complete problem to k-Colorability, we can conclude that k-Colorability is NP-complete as all NP-complete problems are reducible to all other NP-complete problems \cite{gareynp}. We will reduce the well-known 3-SAT problem first to the 3-Colorability problem. The 3-Colorability problem asks if a graph can be colored with 3 (or fewer) colors. Obviously, if a graph can be colored with 2 colors it can also be colored with 3.  From here, the reduction from 3-Colorability to k-Colorability follows.

\subsection{Reduction from 3-SAT to 3-Colorability}
Before we prove Proposition \ref{prop:k-color-np-complete}, we must first introduce 3-SAT.  The 3-SAT problem is an NP-complete problem that is a variation of the original NP-complete problem, boolean satisfiability (SAT). The 3-SAT problem is defined and proven to be NP-complete in \cite{gareynp}. 3-SAT consists of a conjunction of a number of clauses. A clause is a disjunction of a given number of propositions or their negations. We will assume all of our satisfiability problems are in conjunctive normal form (CNF). 3-SAT asks whether the variables can be assigned \emph{true} or \emph{false} in such a way that the whole expression evaluates to \emph{true}. We define SAT and 3-SAT in Definition \ref{def:sat}.

\begin{definition}\label{def:sat}
SAT is a conjunction of $m$ clauses, i.e. $C_1 \wedge C_2 \wedge \dots \wedge C_m$. Each clause is a disjunction of at most $n$ literals, i.e. $L_1 \vee L_2 \vee \dots \vee L_n$. Each literal can be a variable or negative of a variable, i.e. $x_1, \neg x_1, x_2, \neg x_2, \dots , x_k, \neg x_k$ where $k$ is the number of distinct literals. Each literal has a value of $true$ or $false$. 3-SAT is where each clause has exactly 3 literals \((n = 3)\).
\end{definition}

@@

\begin{enumerate}
\item Given an instance of 3-SAT, construct an instance of 3-Colorability
\item Show that if the resulting graph is 3-colorable, then the given 3-SAT problem is satisfied
\end{enumerate}

To show this reduction, we'll create an example 3-SAT problem and reduce it to 3-Colorability. This reduction will create a graph \(G = (V, E)\) having \((2n + 3m + 1)\) vertices and \((3n+6m)\) edges \cite{moret}. The process for creating the graph, described below, was taken from the proof given in \cite{sharma}.

@@

As we see above, \emph{blue} is used for the base vertex and as part of some of the clause triangles. If we let \emph{yellow} denote \emph{true} and \emph{red} denote \emph{false}, we can see that our original 3-SAT is satisfied:
%
\begin{align*}
E &= \left( T \vee \neg T \vee \neg F \right) \wedge \left( \neg T \vee T \vee F \right) \wedge \left(T \vee T \vee F \right) \\
&= \left( T \vee F \vee T \right) \wedge \left( F \vee T \vee F \right) \wedge \left(T \vee T \vee F \right) \\
&= T \wedge T \wedge T \\
&= True
\end{align*}
%
This shows an example of the reduction from 3-SAT to 3-Colorability. If \(G\) was unable to be colored with 3 colors, then we would know that \(E\) is unable to be satisfied. From here, we can reduce 3-Colorability to 4-Colorability, and then generalize that reduction and reduce 4-Colorability to k-Colorability. The proof and illustration of this can be found in \cite{sharma}.

\section{Heuristics}
As we have seen in section \ref{sec:complexity}, solving the vertex coloring problem requires heavy computation. This gives us incentive to use approximation algorithms, called heuristics, to find good, but not necessarily optimal, vertex colorings. We decided to analyze and compare a few of the popular heuristics for solving the VCP. The input for all heuristics we define is a simple graph, $G = (V, E)$.

@@

Heuristic \ref{heuristic-welsh} is the greedy algorithm with degree sequencing. Before applying the greedy algorithm, it orders the vertices according to the decreasing value of their degree. This is also known as the Welsh-Powell algorithm, which is given in \cite{welsh}. It is defined in Algorithm \ref{alg:welsh}.

@@

Heuristic \ref{heuristic-dsatur} is very similar to heuristic \ref{heuristic-welsh}, but colors a graoh based on the \emph{saturation} degree of vertices, which is defined in Definition \ref{def:sat} \cite{spinrad}. The original DSATUR algorithm was proposed by \cite{brelaz}. It is defined in Algorithm \ref{alg:dsatur}.

\begin{definition}[Saturation degree]\label{def:sat}
The saturation degree of a vertex is the number of different colors used for vertices adjacency to it.
\end{definition}

@@

Finding a maximum independent set is NP-hard \cite{karp1972reducibility}, however, meaning we're including a different NP-hard in an already NP-hard problem. We chose to forgo this in favor of just finding a maximal independent set. We define the heuristic in Algorithm \ref{alg:mis}.

@@

\section{Initial Findings}
Before comparing the heuristics, we first wanted to analyze the heuristics individually and look at the solutions they produce.

@@

\begin{prop}
Heuristic \ref{heuristic-greedy} and Heuristic \ref{heuristic-welsh} do not always produce an optimal solution, i.e. they do not always produce a minimum coloring of a graph.
\end{prop}

@@

We can see that $G_1$ is a bipartite graph, defined above by \cite{bondymurty}. Thus, \(G_1\) is 2-colorable as all bipartite graphs are 2-colorable \cite{asratian1998bipartite}. This means \(\chi^{*}(G_1) = 2\). \(G_1\) is an example graph that satisfies our given conditions \((1)\) and \((2)\). The optimal coloring is shown in Figure \ref{fig:ex1-optimal}.

Although Heuristic \ref{heuristic-greedy} and Heuristic \ref{heuristic-welsh} do not always produce an optimal coloring for bipartite graphs, Heuristic \ref{heuristic-dsatur} always will produce an optimal coloring (2 colors). Heuristic \ref{heuristic-dsatur}, however, does not produce optimal colorings for all graphs. Both of these results are shown and proven in \cite{brelaz}. We chose not to create examples of these.

\begin{prop}
Heuristic \ref{heuristic-mis} also does not always produce an optimal solution, i.e. it does not always produce a minimum coloring of a graph.
\end{prop}

To show this, we will create another example that satisfies conditions \((1)\) and \((2)\) from earlier. We've constructed a graph \(G_2 = (V,E)\) for this example. This is shown Figure \ref{fig:ex2-uncolored}.

\begin{figure}[H]
\centering
\includegraphics[scale=0.5]{images/mis-1.png}
\caption{Uncolored original graph \(G_2\)}\label{fig:ex2-uncolored}
\end{figure}

By applying heuristic \ref{heuristic-mis} and finding maximal independent sets, we see that \(G_2\) is \emph{5-colorable}. This is shown Figure \ref{fig:ex2-colored}.

\begin{multicols}{2}
  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.5]{images/mis-2.png}
  \caption{Coloring of \(G_2\) after applying heuristic \ref{heuristic-mis}}\label{fig:ex2-colored}
  \end{figure}

  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.5]{images/mis-3.png}
  \caption{Optimal coloring of \(K\)}\label{fig:ex2-optimal}
  \end{figure}
\end{multicols}

This is a non-optimal solution. We can see that the subgraph created by vertices \(\{5, 6, 7, 8\}\) is complete and thus we need at least 4 colors. This graph is \emph{4-colorable} however, meaning \(\chi^{*}(K) = 4\). Thus, we can see that heuristic \ref{heuristic-mis} does not always give us an optimal solution. The optimal coloring of $G_2$ is shown in Figure \ref{fig:ex2-optimal}.

\begin{prop}
Heuristic \ref{heuristic-greedy} and heuristic \ref{heuristic-welsh} have an upper bound of \(\Delta + 1\).
\end{prop}

To show this, we will create another example that satisfies conditions \((1)\) and \((2)\) from above. After doing some research into bipartite graphs, we learned that \emph{crown graphs} are excellent at showing how bad greedy heuristics can be, as shown and defined in \cite{kordecki}.

@@

\begin{figure}[H]
\centering
\includegraphics[scale=0.38]{images/graph-4.png}
\caption{Uncolored original graph \(G_3\)}
\end{figure}

We can see that \(\Delta(G_3) = 4\). We can also see \(d_{G_3}(v) = 4\) for all \(v \in V\). Thus, in both heuristics \ref{heuristic-greedy} and \ref{heuristic-welsh}, the algorithm would pick an order arbitrarily. We can show using this crown graph the worst-case scenario of these heuristics.

\begin{multicols}{2}
  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.38]{images/graph-5.png}
  \caption{Worst-case coloring of \(G_3\)}\label{fig:ex3-colored}
  \end{figure}

  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.38]{images/graph-6.png}
  \caption{Optimal coloring of \(H\)}\label{fig:ex3-optimal}
  \end{figure}
\end{multicols}

@@

As shown in Figure \ref{fig:ex3-colored}, using either heuristic with this ordering, we get \(\chi(H) = 5\). This gives us \(\frac{|V|}{2}\) colors. This is the worst-case for a crown graph as shown in \cite{johnson}, but this graph also demonstrates the upper bounds for these heuristics.

@@

The proof of Brook's Theorem can be found in \cite{lovasz}. Overall, heuristics can produce undesirable results compared to the optimal solution. In graph \(G_3\), in the worst case, some heuristics produce \(\chi(H) = 5\) when \(\chi^{*}(H) = 2\) as it is a bipartite graph. This is shown in Figure \ref{fig:ex3-optimal}.

@@

\section{Results}
Our main goal was to compare these heuristics to find out which heuristic was most effective in efficiently approximating the vertex coloring problem. we wanted to compare both quality (number of colors used) and efficiency (running time). For this study, we decided to omit the maximal independent set heuristic (heuristic \ref{heuristic-mis}) as it was so different from the other 3 as well as including another NP-hard problem. We implemented the three heuristics in Java and ran simulations according to the following:
\begin{itemize}
  \setlength\itemsep{0.4em}
  \item Graphs were generated based on the \emph{Erdős–Rényi} model \cite{erdds1959random}. This model generates a random graph with approximately $p$ percent of edges.
  \item We analyzed graphs with $10\%, 20\%, \ldots, 90\%$ edges. We called this the type of the graph. This gave us both sparse and dense graphs.
  \item Simulations were run on graphs with 5 different amount of vertices: 50, 100, 250, 500, 1000.
\end{itemize}

@@

\bibliographystyle{abbrv}
\bibliography{references}

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{graphicx}
\usepackage{float}
\usepackage{titling}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[numbib,nottoc]{tocbibind}
\usepackage{enumitem}
\usepackage{multicol}
\usepackage[numbers]{natbib}
\bibliographystyle{abbrvnat}

\setlist[itemize]{topsep=0pt}
\setlist[enumerate]{topsep=0pt}
\setlength\parindent{0pt}
\setlength{\droptitle}{-1.5cm}

@@

\title{\vspace{-0.8cm}Vertex Coloring and Applications}
\date{May 5, 2017}
\author{\Large{Shawn Seymour\vspace{0.2cm}}\\Directed Study (MATH 3993)\\University of Minnesota Morris}

@@

\section{Introduction}
Consider the map of the 48 contiguous states in the USA. Suppose we would like to color each state such that no two states that share a boundary have the same color. How many colors would it take to do this? This problem can be modeled as a \emph{graph}. In general, we could represent every state with a \emph{vertex} and draw an \emph{edge} between two states that share a border. A graph, denoted $G = (V, E)$, is a set of vertices $V$ and a set of edges $E$. A \emph{simple graph} is a loopless, undirected graph where no two edges connect the same pair of vertices. For our purpose, assume all graphs are simple graphs. This problem, along with many others, can be solved using vertex coloring. We can model the map of the 48 contiguous states as a graph. We can represent every state with a vertex and draw an edge between two states that share a border.

\begin{figure}[H]
\centering
\includegraphics[width=0.7\linewidth]{figures/map-colors.pdf}
\caption{A proper vertex coloring of the 48 contiguous states of the USA}\label{fig:map}
\end{figure}

@@

The \emph{chromatic number}, denoted $\chi(G)$, is the minimum number of colors needed to have a proper vertex coloring of a graph $G$. The \emph{vertex coloring problem} (VCP) is to find $\chi(G)$ of any given graph $G$. By applying the vertex coloring problem to our map example, we can determine how many colors one would need to color it with regards to our constraint. An example coloring of our map is shown in Figure \ref{fig:map}. The \emph{k-Colorability problem} asks if a graph can be colored with $k$ colors. If a graph can be colored with $k$ colors, it is said to be \emph{k-colorable}.

@@

In this section, we will first discuss how to classify hard mathematical problems. Then, we will classify the VCP as a mathematically hard problem and show how this classification can be done.

\subsection{Classifying the Vertex Coloring Problem}
We call any problem that can be answered with ``yes'' or ``no'' a \emph{decision problem}. To classify the vertex coloring problem, we must first introduce some classes, defined by \citet{moret}:

@@

Take note that these classes are not exclusive: any member of P is also a member of NP. NP-complete problems can be thought of as the problems that are both in NP and NP-hard. Although it may seem like all problems that are NP are also NP-complete, this is not the case. There are a select few problems in NP that are not NP-complete or in P. See \citet{ladner1975structure}.

@@

It is NP-hard to compute the chromatic number of a graph. This can be proven by showing it is at least as hard as a problem in NP. This is typically done by a reduction (a transformation) to or from 3-SAT, which we'll examine later. This was proven by \citet{garey}.

@@

To prove this, we need to reduce a known NP-complete problem to the vertex coloring problem in polynomial time. By reducing a known NP-complete problem to k-Colorability, we can conclude that k-Colorability is NP-complete as all NP-complete problems are reducible to all other NP-complete problems. See \citet{gareynp}. We will reduce the well-known 3-SAT problem first to the 3-Colorability problem. The 3-Colorability problem asks if a graph can be colored with 3 (or fewer) colors. Obviously, if a graph can be colored with 2 colors it can also be colored with 3.  From here, the reduction from 3-Colorability to k-Colorability follows.

\subsection{Reduction from 3-SAT to 3-Colorability}
Before we prove Proposition \ref{prop:k-color-np-complete}, we must first introduce 3-SAT.  The 3-SAT problem is a decision problem that is a variation of the original NP-complete problem, Boolean satisfiability (SAT). The 3-SAT problem was proven to be NP-complete. See \citet{gareynp}. SAT consists of a conjunction of $m$ clauses, i.e. $C_1 \wedge C_2 \wedge \dots \wedge C_m$. Each clause is a disjunction of at most $n$ literals, i.e. $L_1 \vee L_2 \vee \dots \vee L_n$. Each literal can be a variable or negative of a variable, i.e. $x_1, \neg x_1, x_2, \neg x_2, \dots , x_k, \neg x_k$ where $k$ is the number of distinct literals. Each literal has a value of $true$ or $false$. 3-SAT has exactly 3 literals in each clause $(n = 3)$. We will assume all of our satisfiability problems are in conjunctive normal form (CNF). 3-SAT asks whether the variables can be assigned \emph{true} or \emph{false} in such a way that the whole expression evaluates to \emph{true}. We define SAT and 3-SAT in Definition \ref{def:sat}.

\begin{definition}\label{def:sat}
SAT is a decision problem that decides the truth value of a Boolean function based on the evaluation of the clause. It determines whether there are truth values to each clause in a set of clauses that will return a truth value to the given Boolean function. 3-SAT is a variation of SAT where each clause contains exactly 3 literals.
\end{definition}

@@

\begin{enumerate}
\item Construct an instance of 3-Colorability, given an instance of 3-SAT
\item Show that if the resulting graph is 3-colorable, then the given 3-SAT problem is satisfied
\end{enumerate}

To show this reduction, we'll create an example 3-SAT problem and reduce it to 3-Colorability. This reduction will create a graph \(G = (V, E)\) having \((2n + 3m + 1)\) vertices and \((3n+6m)\) edges. See \citet{moret}. The process for creating the graph, described below, was taken from the proof given by \citet{sharma}.

@@

As we see above, \emph{blue} is used for the base vertex and as part of some of the clause triangles. If we let \emph{yellow} denote \emph{true} and \emph{red} denote \emph{false}, we can see that our original 3-SAT is satisfied:
%
\begin{align*}
E &= \left( T \vee \neg T \vee \neg F \right) \wedge \left( \neg T \vee T \vee F \right) \wedge \left(T \vee T \vee F \right) \\
&= \left( T \vee F \vee T \right) \wedge \left( F \vee T \vee F \right) \wedge \left(T \vee T \vee F \right) \\
&= T \wedge T \wedge T \\
&= True
\end{align*}
%
This shows an example of the reduction from 3-SAT to 3-Colorability. If \(G\) was unable to be colored with 3 colors, then we would know that \(E\) is unable to be satisfied. From here, we can reduce 3-Colorability to 4-Colorability, and then generalize that reduction and reduce 4-Colorability to k-Colorability. The proof and illustration of this can be found in \citet{sharma}.

\section{Heuristics}
As we have seen in section \ref{sec:complexity}, the vertex coloring problem is NP-hard and thus solving it to optimality for any given instance of a graph is computationally ``impossible''. This gives us incentive to use approximation algorithms, called heuristics, to find good, but not necessarily optimal, vertex colorings. We will analyze and compare a few of the popular heuristics for solving the VCP. However, in this section, we will describe the four heuristics to solve the VCP. The input for all heuristics we define is a simple graph, $G = (V, E)$.

@@

Heuristic \ref{heuristic-welsh} is the greedy algorithm with degree sequencing. Before applying the greedy algorithm, it orders the vertices according to the decreasing value of their degree. This is also known as the Welsh-Powell algorithm, which was defined by \citet{welsh}. It is defined in Algorithm \ref{alg:welsh}.

@@

Heuristic \ref{heuristic-dsatur} is very similar to heuristic \ref{heuristic-welsh}, but colors a graoh based on the \emph{saturation} degree of vertices, which is defined in Definition \ref{def:saturation} \cite{spinrad}. The original DSATUR algorithm was proposed by \citet{brelaz}. It is defined in Algorithm \ref{alg:dsatur}.

\begin{definition}[Saturation degree]\label{def:saturation}
The saturation degree of a vertex is the number of different colors used for vertices adjacency to it.
\end{definition}

@@

Finding a maximum independent set is NP-hard as detailed by \citet{karp1972reducibility}. This means we're including a different NP-hard in an already NP-hard problem. We chose to forgo this in favor of just finding a maximal independent set. We define the heuristic in Algorithm \ref{alg:mis}.

@@

\section{Analysis of Heuristics A - D}
Before comparing the heuristics, we first wanted to analyze the heuristics individually and look at the solutions they produce.

@@

\begin{prop}
Heuristic \ref{heuristic-greedy} and Heuristic \ref{heuristic-welsh} do not produce optimal solutions, i.e. they do not always produce a minimum coloring for all graphs.
\end{prop}

@@

We can see that $G_1$ is a bipartite graph. Thus, \(G_1\) is 2-colorable as all bipartite graphs are 2-colorable. See \citet{asratian1998bipartite}. This means \(\chi^{*}(G_1) = 2\). \(G_1\) is an example graph that satisfies our given conditions \((1)\) and \((2)\). The optimal coloring is shown in Figure \ref{fig:ex1-optimal}.

Although Heuristic \ref{heuristic-greedy} and Heuristic \ref{heuristic-welsh} do not always produce an optimal coloring for bipartite graphs, Heuristic \ref{heuristic-dsatur} always will produce an optimal coloring (2 colors). Heuristic \ref{heuristic-dsatur}, however, does not produce optimal colorings for all graphs. Both of these results are shown and proven by \citet{brelaz}.

\begin{prop}
Heuristic \ref{heuristic-dsatur} does not produce an optimal solution, i.e. it does not always produce a minimum coloring for all graphs.
\end{prop}

To show this, we've created a graph $G_2 = (V, E)$ for this example. This is shown in.

\begin{figure}[H]
\centering
\includegraphics[scale=0.5]{images/dsatur-1.png}
\caption{Uncolored original graph \(G_2\)}\label{fig:ex2-uncolored}
\end{figure}

By applying heuristic \ref{heuristic-dsatur}, we can see that $G_2$ is \emph{4-colorable}. This is shown in Figure \ref{fig:ex2-colored}.

\begin{multicols}{2}
  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.5]{images/dsatur-2.png}
  \caption{Coloring of \(G_2\) after applying heuristic \ref{heuristic-dsatur}}\label{fig:ex2-colored}
  \end{figure}

  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.5]{images/dsatur-3.png}
  \caption{Optimal coloring of \(G_2\)}\label{fig:ex2-optimal}
  \end{figure}
\end{multicols}

This is a non-optimal solution as this graph is \emph{3-colorable}. This means \(\chi^{*}(G_2) = 3\). Thus, we can see that heuristic \ref{heuristic-dsatur} does not always give us an optimal solution. The optimal coloring of $G_2$ is shown in Figure \ref{fig:ex2-optimal}.

\begin{prop}
Heuristic \ref{heuristic-mis} does not produce an optimal solution, i.e. it does not always produce a minimum coloring for all graphs.
\end{prop}

To show this, we will create another example that satisfies conditions \((1)\) and \((2)\) from earlier. We've constructed a graph \(G_3 = (V,E)\) for this example. This is shown Figure \ref{fig:ex3-uncolored}.

\begin{figure}[H]
\centering
\includegraphics[scale=0.5]{images/mis-1.png}
\caption{Uncolored original graph \(G_3\)}\label{fig:ex3-uncolored}
\end{figure}

By applying heuristic \ref{heuristic-mis} and finding maximal independent sets, we see that \(G_3\) is \emph{5-colorable}. This is shown Figure \ref{fig:ex3-colored}.

\begin{multicols}{2}
  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.5]{images/mis-2.png}
  \caption{Coloring of \(G_3\) after applying heuristic \ref{heuristic-mis}}\label{fig:ex3-colored}
  \end{figure}

  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.5]{images/mis-3.png}
  \caption{Optimal coloring of \(G_3\)}\label{fig:ex3-optimal}
  \end{figure}
\end{multicols}

This is a non-optimal solution. We can see that the subgraph created by vertices \(\{5, 6, 7, 8\}\) is complete and thus we need at least 4 colors. This graph is \emph{4-colorable} however, meaning \(\chi^{*}(G_3) = 4\). Thus, we can see that heuristic \ref{heuristic-mis} does not always give us an optimal solution. The optimal coloring of $G_3$ is shown in Figure \ref{fig:ex3-optimal}.

\begin{prop}
Heuristic \ref{heuristic-greedy} and heuristic \ref{heuristic-welsh} yield an upper bound of \(\Delta + 1\).
\end{prop}

To show this, we will create another example that satisfies conditions \((1)\) and \((2)\) from above. After doing some research into bipartite graphs, we learned that \emph{crown graphs} are excellent at showing how bad greedy heuristics can be, as shown and defined by \citet{kordecki}.

@@

\begin{figure}[H]
\centering
\includegraphics[scale=0.38]{images/graph-4.png}
\caption{Uncolored original graph \(G_4\)}
\end{figure}

We can see that \(\Delta(G_4) = 4\). We can also see \(d_{G_4}(v) = 4\) for all \(v \in V\). Thus, in both heuristics \ref{heuristic-greedy} and \ref{heuristic-welsh}, the algorithm would pick an order arbitrarily. We can show using this crown graph the worst-case scenario of these heuristics.

\begin{multicols}{2}
  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.38]{images/graph-5.png}
  \caption{Worst-case coloring of \(G_4\)}\label{fig:ex4-colored}
  \end{figure}

  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.38]{images/graph-6.png}
  \caption{Optimal coloring of \(G_4\)}\label{fig:ex4-optimal}
  \end{figure}
\end{multicols}

@@

As shown in Figure \ref{fig:ex4-colored}, using either heuristic with this ordering, we get \(\chi(G_4) = 5\). This gives us \(\frac{|V|}{2}\) colors. This is the worst-case for a crown graph. See \citet{johnson}. This graph also demonstrates the upper bounds for these heuristics.

@@

The proof of Brook's Theorem is shown by \citet{lovasz}. Overall, heuristics can produce undesirable results compared to the optimal solution. In graph \(G_4\), in the worst case, some heuristics produce \(\chi(G_4) = 5\) when \(\chi^{*}(G_4) = 2\) as it is a bipartite graph. This is shown in Figure \ref{fig:ex4-optimal}.

@@

\section{Results}
Our main goal was to compare these heuristics to find out which heuristic was most effective in efficiently approximating the vertex coloring problem. we wanted to compare both quality (number of colors used) and efficiency (running time). For this study, we decided to omit the maximal independent set heuristic (heuristic \ref{heuristic-mis}) as it was so different from the other 3 as well as including another NP-hard problem. We implemented the three heuristics in Java and ran simulations according to the following:
\begin{itemize}
  \setlength\itemsep{0.4em}
  \item Graphs were randomly generated based on the \emph{Erdős–Rényi} model, given in \citet{erdds1959random}. This model generates a random graph with approximately $p$ percent of edges.
  \item We analyzed graphs with $10\%, 20\%, \ldots, 90\%$ edges. We called this the type of the graph. This gave us both sparse and dense graphs.
  \item Simulations were run on graphs with 5 different amount of vertices: 50, 100, 250, 500, 1000.
\end{itemize}

@@

\bibliography{references}
